---
output:
  pdf_document: default
  html_document: default
---
# Part III: Coding 
  
1. You are working on the following dataframe:

```{r, include = FALSE}
library(tidyverse)
library(Biostrings)
library(ape)
library(dplyr)
```

```{r, eval=TRUE}
student_df <- data.frame(
  Name = c("Alice", "Bob", "Charlie", "Deborah", "Eliza", "Frank", "Gigi", "Henry"),
  Major = c("Bio", "CompSci", "Neuro", "Bio", "Bio", "CompSci", "Neuro", "Bio"),
  Minor = c(NA, NA, "Data Analytics", NA, "Chem", NA, "Bio", NA),
  GPA = c(2.6, 3.4, 3.7, 3.8, 2.9, 3.0, 3.5, 2.6)
)
print(student_df)
```

(a) What should you get when you run `student_df$Name[8]`?

(b) What would you expect the output of the following code to be?  You can describe it or make the output yourself.

```{r, eval=FALSE}
student_df %>%
  filter(Minor != NA)
```

\newpage

You wish to depict the average GPA for each student based on major:

```{r, eval=TRUE}
GPA_by_major <- ggplot(student_df, aes(x=Major, y = GPA))
GPA_by_major + geom_boxplot(aes(color=Major)) 
```

(c) Circle the part of this code which makes the boxplots for Biology, Computer Science, and Neuroscience each a different color.

(d) Let's say you changed the second line of code to `GPA_by_major + geom_boxplot(aes(color=Major)) + labs(y = "GPA at the End of Freshman year")`.  What effect would this have on the plot?

\newpage

2. Consider the following function, which takes an aligned fasta file, takes a subset of the nucleic acids, and makes a phylogeny of the fasta file:
  
```{r, eval=TRUE}
phylogeny_from_fasta = function(input_fasta, start, end){
  seqs <- readDNAStringSet(input_fasta)
  if(length(seqs) > 10){
    seqs = seqs[1:10]
  }
  sub_seqs <- subseq(seqs, start = start, end = end)
  dist_matrix <- pwalign::stringDist(sub_seqs, method = "hamming") 
  nj_tree <- nj(dist_matrix)
  plot(nj_tree, main="Neighbor-Joining Tree") 
}
phylogeny_from_fasta("Full_cmv_259sequences_Aln.fasta", 30000, 31000)
```

(a) Circle the line of code that specifies the region of DNA for which to make a phylogeny (in the example, from nucleotides 30000 to 31000).

(b) What is the purpose of lines 3-5 (`if(length(seqs) > 10){seqs = seqs[1:10]}`)?

